Nilpotent Invariants \\ for Generic Discrete Series of Real Groups

### 2024-10-02 whittaker.tex
Affiliations and typography

%edited starting 8/27/17 for re-submission to Duke
\documentclass[10pt,leqno]{article}
\usepackage{verbatim}
\usepackage{amssymb, amscd}
\usepackage{mathtools}
%\usepackage{amsrefs}
\usepackage{rotating}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[colorlinks=true, linktoc=page, citecolor=blue, linkcolor=blue, urlcolor=blue]{hyperref}

%\usepackage{showkeys}
\usepackage{tabularx}
\setlength\extrarowheight{4pt}   %spacing in tables
\usepackage{theorem}
\usepackage[matrix,tips,frame,color,line,poly,curve]{xy}
\renewcommand{\labelenumi}{(\arabic{enumi})}
\newcommand\kappaarrow[2]{#1\overset\kappa\rightarrow#2}
\newtheorem{theorem}[equation]{Theorem}
\newtheorem{corollary}[equation]{Corollary}
\newtheorem{definition}[equation]{Definition}
\newtheorem{lemma}[equation]{Lemma}
\newtheorem{desideratum}[equation]{Desideratum}
\newtheorem{conjecture}[equation]{Conjecture}
\newtheorem{proposition}[equation]{Proposition}
\newtheorem{remark}[equation]{Remark}
{\theorembodyfont{\rmfamily}\newtheorem{theoremplain}[equation]{Theorem}
\newtheorem{remarkplain}[equation]{Remark}
\newtheorem{editorialremarkplain}[equation]{Editorial Remark}
\newtheorem{exampleplain}[equation]{Example}
\newtheorem{corollaryplain}[equation]{Corollary}
\newtheorem{mytable}[equation]{Table}
}

@@

\renewcommand{\t}{\mathfrak t}
\newcommand{\su}{\mathfrak s\mathfrak u}
\newcommand{\g}{\mathfrak g}
\newcommand{\gder}{\mathfrak g_{\mathrm{der}}}
\newcommand\inv{^{-1}}
\newcommand\wh{\widehat}
\newcommand{\GL}{\text{GL}}
\newcommand{\SL}{\text{SL}}
\newcommand{\SO}{\text{SO}}
\newcommand{\SU}{\text{SU}}
\newcommand{\Spin}{\text{Spin}}
\newcommand{\chGGamma}{\phantom{a}^\vee G^\Gamma}
\newcommand{\GGamma}{G^\Gamma}
\newcommand{\s}{\mathfrak s}
\newcommand{\w}{\mathfrak w}
\newcommand{\AV}{\mathrm{AV}}
\newcommand{\KS}{\mathrm{KS}}
\newcommand{\Wh}{\mathrm{Wh}}
\newcommand{\WF}{\mathrm{WF}}
\newcommand{\AC}{\mathrm{AC}}
\newcommand{\AVann}{\mathrm{AV}_{\mathrm{ann}}}
\newcommand{\GK}{\mathrm{GK}}
\newcommand{\Op}{\O_p}
\newcommand{\Kostant}[1]{\mathcal{K}(#1)}
\newcommand{\ECom}{\mathcal{E}^{\mathcal{C}}_\Omega}
\begin{document}
%Jeffrey Adams University of Maryland/IDA Center for Computing Sciences
%jda@math.umd.edu
\title{Nilpotent Invariants \\ for Generic Discrete Series of Real Groups}
\author{Jeffrey Adams \& Alexandre Afgoustidis}
\maketitle

@@

The set $\Wh(\pi)$ is
a useful invariant of generic representations.
For instance, in the local Langlands correspondence, if $\Pi$ is a tempered $L$-packet for $G(\R)$ and $\w$ is a Whittaker datum, then $\Pi$ contains a unique representation
having a $\w$-Whittaker model. This plays a role in describing the internal structure of $L$-packets.
If $\Pi$ is a discrete series $L$-packet then each generic representation in the packet has a unique
Whittaker model, so the generic representations in $\Pi$ are parametrized by their Whittaker models.
It is conjectured that these properties hold in the $p$-adic case as well, and this is known in many cases.

There are other invariants attached to an
irreducible representation $\pi$.  We will be concerned with two of
them, defined very differently. The first is the \emph{wavefront set} $\WF(\pi)$,
which is an analytic invariant.
The second is the \emph{associated variety} $\AV(\pi)$, which is
defined  algebraically.
Both consist of nilpotent elements of the dual of the Lie algebra of~$G$.
The invariants $\Wh(\pi), \WF(\pi)$ and $\AV(\pi)$ are closely related to each other.

@@

Let~$\sigma$ be the anti-holomorphic involution defining~$G(\R)$.
Set $\g=\Lie(G)$ and $\g(\R)=\Lie(G(\R))$. 
Choose a Cartan involution $\theta$ for $G$. This means:
$\theta$ is an algebraic involution, commuting with~$\sigma$,
and $G(\R)^\theta$ is a maximal compact subgroup of~$G(\R)$. Set $K=G^\theta$ and $K(\R)=G(\R)^\theta=K^\sigma$.
Let $\s$ be the $-1$ eigenspace of $\theta$ acting on~$\g$, where we use the notation $\sigma, \theta$ also for the automorphisms of~$\g$ induced by the above involutions.

@@

A related invariant is the associated variety of the annihilator $I_\pi$ of $\pi$ in the universal enveloping algebra, 
denoted $\AVann(\pi)$.
This is a subset of $\N^*\subset\g^*$, and is the closure of a single complex nilpotent orbit: see \cite[Section 1, Theorem 4.7]{vogan_bowdoin}. Both $\AV(\pi)$ and $\AVann(\pi)$ depend only on the equivalence class of~$\pi$.

@@

Let us first set up some general notation. When~$A$ is a group and $\tau$ is an involutive automorphism of~$A$, consider the cohomology sets $H^0(\tau, A)$ and $H^1(\tau, A)$ attached to the action of $\Z/2\Z$ on~$A$ by~$\tau$. We shall view   $H^0(\tau, A)$ as the fixed-point-set $A^{\tau}$, and $H^1(\tau,A)$ as the quotient of $A^{-\tau} = \{ a \in A, \tau(a)=a^{-1}\}$ by the equivalence relation $a \sim x a \tau(x^{-1})$ for all $x \in A$. The set $H^1(\tau, A)$ has a distinguished point $1$, coming from the identity of~$A$; but in general it is just a pointed set, and not a group if~$A$ is nonabelian.

@@

Because of Lemmas~\ref{lem:iso_q} and \ref{lem:action_q}, we write $Q(G)$ for the group $Q_\sigma(G)$ or $Q_\theta(G)$, depending on the situation.

@@

In this section only we need to  choose a particular an isomorphism $\gder\simeq \gder^*$.
For this we let $\kappa$ be the killing form, and define $\psi_\kappa:\gder\rightarrow \gder^*$ by
$$
\psi_\kappa(X)(Y)=\kappa(X,Y)\quad (X,Y\in \gder).
$$

@@

\begin{proof}
This follows from Lemma \ref{l:pi_to_av} for the associated variety;
from Lemma \ref{lem:WF_and_AC} for the wave front set, and Lemma \ref{lem:Kost_and_AC} (or Corollary \ref{cor:matumoto})
for the Whittaker model.
\end{proof}

@@

Furthermore $E_\R,F_\R$ are in the $-1$-eigenspace of $\sigma$, and we have
$\theta(E_\R)=F_\R$ and $\theta(H_\R)=-H_\R$.

@@

Set $g=\frac 1{\sqrt 2}\begin{pmatrix}1&i\\i&1
\end{pmatrix}$, and take $E_\theta,F_\theta,H_\theta$ be the conjugates of $E_\R,F_\R,H_\R$ by~$g$:
$$
E_\theta=\frac12\begin{pmatrix}1&i\\i&-1\end{pmatrix}, \quad
F_\theta=\frac12\begin{pmatrix}1&-i\\-i&-1\end{pmatrix},\quad
H_\theta=\begin{pmatrix}0&-i\\i&0\end{pmatrix}.
$$
These also satisfy the standard $SL(2)$ commutation relations; $E_\theta,F_\theta$ are in the $-1$-eigenspace of $\theta$; 
and  $\sigma(E_\theta)=F_\theta,\sigma(H_\theta)=-H_\theta$.

>>>>

\documentclass[10pt,leqno]{article}

\usepackage{verbatim}
\usepackage{amssymb, amscd}
\usepackage{mathtools}
\usepackage{rotating}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage[colorlinks=true, linktoc=page, citecolor=blue, linkcolor=blue, urlcolor=blue]{hyperref}

\usepackage{tabularx}
\setlength\extrarowheight{4pt}   %spacing in tables
\usepackage{theorem}
\usepackage[matrix,tips,frame,color,line,poly,curve]{xy}
\renewcommand{\labelenumi}{(\arabic{enumi})}
\newcommand\kappaarrow[2]{#1\overset\kappa\rightarrow#2}
\newtheorem{theorem}[equation]{Theorem}
\newtheorem{corollary}[equation]{Corollary}
\newtheorem{definition}[equation]{Definition}
\newtheorem{lemma}[equation]{Lemma}
\newtheorem{desideratum}[equation]{Desideratum}
\newtheorem{conjecture}[equation]{Conjecture}
\newtheorem{proposition}[equation]{Proposition}
\newtheorem{remark}[equation]{Remark}
{\theorembodyfont{\rmfamily}\newtheorem{theoremplain}[equation]{Theorem}
\newtheorem{remarkplain}[equation]{Remark}
\newtheorem{editorialremarkplain}[equation]{Editorial Remark}
\newtheorem{exampleplain}[equation]{Example}
\newtheorem{corollaryplain}[equation]{Corollary}
\newtheorem{mytable}[equation]{Table}
}

@@

\renewcommand{\t}{\mathfrak t}
\newcommand{\su}{\mathfrak s\mathfrak u}
\newcommand{\g}{\mathfrak g}
\newcommand{\gder}{\mathfrak g_{\mathrm{der}}}
\newcommand\inv{^{-1}}
\newcommand\wh{\widehat}
\newcommand{\GL}{\text{GL}}
\newcommand{\SL}{\text{SL}}
\newcommand{\SO}{\text{SO}}
\newcommand{\SU}{\text{SU}}
\newcommand{\Spin}{\text{Spin}}
\newcommand{\chGGamma}{\phantom{a}^\vee G^\Gamma}
\newcommand{\GGamma}{G^\Gamma}
\newcommand{\s}{\mathfrak s}
\newcommand{\w}{\mathfrak w}
\newcommand{\AV}{\mathrm{AV}}
\newcommand{\KS}{\mathrm{KS}}
\newcommand{\Wh}{\mathrm{Wh}}
\newcommand{\WF}{\mathrm{WF}}
\newcommand{\AC}{\mathrm{AC}}
\newcommand{\AVann}{\mathrm{AV}_{\mathrm{ann}}}
\newcommand{\GK}{\mathrm{GK}}
\newcommand{\Op}{\O_p}
\newcommand{\Kostant}[1]{\mathcal{K}(#1)}
\newcommand{\ECom}{\mathcal{E}^{\mathcal{C}}_\Omega}
\begin{document}

\title{Nilpotent Invariants \\ for Generic Discrete Series of Real Groups}
\author{Jeffrey Adams\footnote{University of Maryland \& IDA Center for Computing Sciences ; \url{jda@math.umd.edu}} {} \& Alexandre Afgoustidis\footnote{CNRS \& Université de Lorraine, Nancy \& Metz ; \url{alexandre.afgoustidis@math.cnrs.fr}}}
%\email{alexandre.afgoustidis@math.cnrs.fr}

\maketitle

@@

The set $\Wh(\pi)$ is
a useful invariant of generic representations.
For instance, in the local Langlands correspondence, if $\Pi$ is a tempered $L$-packet for $G(\R)$ and $\w$ is a Whittaker datum, then $\Pi$ contains a unique representation
having a $\w$-Whittaker model. This plays a role in describing the internal structure of $L$-packets.
If $\Pi$ is a discrete series $L$-packet then each generic representation in the packet has a unique
Whittaker model, so the generic representations in~$\Pi$ are parametrized by their Whittaker models.
It is conjectured that these properties hold in the $p$-adic case as well, and this is known in many cases.

There are other invariants attached to an
irreducible representation $\pi$.  We will be concerned with two of
them, defined very differently. The first is the \emph{wavefront set} $\WF(\pi)$,
which is an analytic invariant.
The second is the \emph{associated variety} $\AV(\pi)$, which is
defined  algebraically.
Both consist of nilpotent elements of the dual of the Lie algebra of~$G$.
The invariants $\Wh(\pi), \WF(\pi)$ and~$\AV(\pi)$ are closely related to each other.

@@

Let~$\sigma$ be the anti-holomorphic involution defining~$G(\R)$.
Set $\g=\Lie(G)$ and $\g(\R)=\Lie(G(\R))$. 
Choose a Cartan involution $\theta$ for $G$. This means:
$\theta$ is an algebraic involution, commuting with~$\sigma$,
and $G(\R)^\theta$ is a maximal compact subgroup of~$G(\R)$. Set $K=G^\theta$ and $K(\R)=G(\R)^\theta=K^\sigma$.
Let $\s$ be the $(-1)$-eigenspace of $\theta$ acting on~$\g$, where we use the notation $\sigma, \theta$ also for the automorphisms of~$\g$ induced by the above involutions.

@@

A related invariant is the associated variety of the annihilator $I_\pi$ of $\pi$ in the universal enveloping algebra, 
denoted $\AVann(\pi)$.
This is a subset of $\N^*$ of~$\g^*$, and is the closure of a single complex nilpotent orbit: see \cite[Section 1, Theorem~4.7]{vogan_bowdoin}. Both $\AV(\pi)$ and $\AVann(\pi)$ depend only on the equivalence class of~$\pi$.

@@

Let us first set up some general notation. When~$A$ is a group and $\tau$ is an involutive automorphism of~$A$, consider the cohomology sets $H^0(\tau, A)$ and $H^1(\tau, A)$ attached to the action of $\Z/2\Z$ on~$A$ by~$\tau$. We shall view   $H^0(\tau, A)$ as the fixed-point-set $A^{\tau}$, and $H^1(\tau,A)$ as the quotient of \mbox{$A^{-\tau} = \{ a \in A, \tau(a)=a^{-1}\}$} by the equivalence relation $a \sim x a \tau(x^{-1})$ for all $x \in A$. The set $H^1(\tau, A)$ has a distinguished point $1$, coming from the identity of~$A$; but in general it is just a pointed set, and not a group if~$A$ is nonabelian.

@@

Because of Lemmas~\ref{lem:iso_q} and \ref{lem:action_q}, we write $Q(G)$ for the group $Q_\sigma(G)$ or~$Q_\theta(G)$, depending on the situation.

@@

In this section only we need to  choose a particular an isomorphism \mbox{$\gder\simeq \gder^*$}.
For this we let $\kappa$ be the killing form, and define $\psi_\kappa:\gder\rightarrow \gder^*$ by
$$
\psi_\kappa(X)(Y)=\kappa(X,Y)\quad (X,Y\in \gder).
$$

@@

\begin{proof}
This follows from Lemma \ref{l:pi_to_av} for the associated variety;
from Lemma~\ref{lem:WF_and_AC} and Proposition~\ref{prop:AC_chamber} for the wave front set, and Lemma \ref{lem:Kost_and_AC} (or Corollary \ref{cor:matumoto})
for the Whittaker model.
\end{proof}

@@

Furthermore $E_\R,F_\R$ are in the $(-1)$-eigenspace of $\sigma$, and we have
$\theta(E_\R)=F_\R$ and $\theta(H_\R)=-H_\R$.

@@

Set $g=\frac 1{\sqrt 2}\begin{pmatrix}1&i\\i&1
\end{pmatrix}$, and take $E_\theta,F_\theta,H_\theta$ be the conjugates of $E_\R,F_\R,H_\R$ by~$g$:
$$
E_\theta=\frac12\begin{pmatrix}1&i\\i&-1\end{pmatrix}, \quad
F_\theta=\frac12\begin{pmatrix}1&-i\\-i&-1\end{pmatrix},\quad
H_\theta=\begin{pmatrix}0&-i\\i&0\end{pmatrix}.
$$
These also satisfy the standard $SL(2)$ commutation relations; $E_\theta,F_\theta$ are in the $(-1)$-eigenspace of $\theta$; 
and  $\sigma(E_\theta)=F_\theta,\sigma(H_\theta)=-H_\theta$.
